\section{High Availability}


High Availability is usually implemented by duplicating the service over several distinct physical or
virtual hosts. The hosts are called a \"cluster\". The software on the hosts and other resources
that are duplicated and necessary for the service are managed by what is called a \"cluster manager\".
Said manager handles cluster communication and failover action over several hosts if a local service,
that is protected by HA\footnote{High Availability} goes down.

\section{Cluster Types}

There are diverse cluster types for different purposes.
\begin{itemize}
\item High Availability Cluster to increase the availability of services.
\item High Performance Clusters to increase the performance of services.
\item Load Balancing Cluster to balance the load of a service.
\end{itemize}

\section{Data Storage}

The hosts in the cluster usually provide a service for which persistent data storage is necessary.
There are two different feasible ways to implement this in a cluster environment:
\begin{itemize}
\item Using a storage area network
\item Using a network access storage
\end{itemize}

Using a storage area network in a cluster environment is good choice as it provides highly available access if the SAN\footnote{storage area network} is layed out for high availability. Usually 
Using network access storage is a more common solution for providing access to data in an environment where concurrent access is needed, but cluster fiel systems are not an option.

In this work, we will only look at High Availability Clusters.

\section{Software}
To implement High Availability, different types of software for different purposes are used.
As mentioned before, a cluster manager is used for communication between the different hosts that
provide services in the cluster. Services are commonly called "resources" in the cluster world.
Another software is the resource manager, which manages the resources themselves.
This includes starting, stopping and monitoring the resources.
\subsection{Pacemaker}
\subsection{Corosync}
